---
title: "Class 11 Lab Session"
author: 'Chloe Wong (PID: A16893383)'
date: "2024-11-07"
output: pdf_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Section 1. Proportion of G/G in a population

Downloaded a CSV file from Ensemble  <
https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```


## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in  assessing genetic differences on a population scale. 

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

Lets make a boxplot

```{r}
plot <- ggplot(expr) + aes(x=geno, exp, fill=geno) +
  geom_boxplot(notch=TRUE)
plot
```

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

(Code above for sample sizes)

The sample size for A/A is 108. The sample size for A/G is 233. The sample size for G/G is 121. The median expression for A/A is 30.779. The median for A/G is 25.640. The median for G/G is 20.004.

```{r}
summary(expr)
```


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

In the boxplot, the expression value of A/A is significantly higher than the expression value of G/G. The SNP affects the expression of ORMDL3 as the expression is seen to be approximately 10% greater in A/A compared to G/G.

