---
title: "Class 10: Halloween Mini-Project"
author: "Chloe Wong (PID: A16893383)"
format: pdf
---

Today is Halloween, an ole Irish holiday, let's celebrate by eating candy. 

We will explore some data all about Halloween candy from the 538 website. 

```{r}
candy_file <- "candy-data.csv"

candy = read.csv("candy-data.csv", row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```


```{r}
rownames(candy)
```


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

```{r}
sum(candy$chocolate)
```

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Twix", ]$winpercent
```

```{r}
candy["Skittles original", "winpercent"]
```

```{r}
candy["Rolo", "winpercent"]
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library(dplyr)

candy %>%
  filter(rownames(candy)=="Haribo Happy Cola") %>%
  select(winpercent)
```

> Q. Find fruity candy with a win percent above 50%

```{r}
candy %>%
  filter(winpercent >50) %>%
  filter(fruity==1)
  
```

```{r}
top.candy <- candy[candy$winpercent > 50,]
top.candy[top.candy$fruity==1,]
```

To get a quick insight into a new dataset some folks like using the skimer package and its `skim()` function. 

```{r}
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Looks like the `winpercent` variable or column is measured on a different scale than everything else! I will need to scale my data before doing any analysis like PCA etc. 

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero means it is not chocolate (false) and a 1 means it is chocolate (true) for the candy$chocolate column.

> Q8. Plot a histogram of winpercent values

We can do this a few ways, e.g. the "base" R `hist()` function or with `ggplot()`

```{r}
hist(candy$winpercent, breaks=100)
```

```{r}
library(ggplot2)

ggplot(candy) +
  aes(winpercent) +
  geom_histogram(binwidth=8) +
  theme_bw()
```

> Q9. Is the distribution of winpercent values symmetrical?

No, the distribution of winpercent values are not symmetrical. It looks like it is slanted towards the left side. 

> Q10. Is the center of the distribution above or below 50%?

The center of the distribution is below 50%. It is at 47.83%.

```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

On average, chocolate candy (60.92) is higher ranked than fruit candy (44.12%).

```{r}
fruit.candy <- candy |>
  filter(fruity==1)

summary(fruit.candy$winpercent)
```

```{r}
summary(candy[as.logical(candy$chocolate),]$winpercent)
```

```{r}
choc.candy <- candy |>
  filter(chocolate==1)

summary(choc.candy$winpercent)
```

> Q12. Is this difference statistically significant?

Yes, the difference is statistically significant as the p-value is extremely small (2.871e-08).

```{r}
t.test(choc.candy$winpercent, fruit.candy$winpercent)
```

> Q13. What are the five least liked candy types in this set?

The five least liked candy types in this set are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters. 

```{r}
play <- c("d","a","c")
sort(play)
order(play)

play[ order(play) ]
```

```{r}
head(candy[order( candy$winpercent ),], 5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

The top 5 all time favorite candy types out of this set are Snickers, Kit Kat, Twix, Reese's Miniatures, and Reese's Peanut Butter Cup. 

```{r}
tail(candy[order( candy$winpercent ),], 5)
```

Let's do a barplot of winpercent values 

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()

```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(x=winpercent, 
      y=reorder(rownames(candy), winpercent),
      fill=chocolate) +

  geom_col()
```

I want a more custom color scheme where I can see both chocolate and bar and fruity etc. ll from the one plot. To do this we can roll our own color vector... 

```{r}
# Place holder color vector
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

```

```{r}
# Use blue for your favorite candy!
my_cols[ rownames(candy)=="Twix"] <- "blue"
#mycols
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets (shown from graph above).

> Q18. What is the best ranked fruity candy?

Starburst (shown from graph above).

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Minatures. 

Plot of winpercent vs pricepercent to see what would be the best candy to buy... 

```{r}
my_cols[as.logical(candy$fruity)] = "red"
```

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent) +
  geom_point(col=my_cols)
```

Add labels

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text(col=my_cols)
```

Make the labels non-overlapping

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols, max.overlaps=8)
```

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The top 5 most expensive candy types in the dataset are Nik L Nip, Ring Pop, Nestle Smarties, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular out of these are Nik L Nip. 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```


```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij, diag = F, type = "upper")
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The two variables that are anti-correlated are chocolate and fruity.

> Q23. Similarly, what two variables are most positively correlated?

The two variables that are the most positively correlated are chocolate and bar. 


#Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], col=my_cols, pch=16)
```

How do the original variables (columns) contribute to the new PCs. I will look at PC1 here. 

```{r}
pca$rotation
```

```{r}
loadings <- as.data.frame(pca$rotation)

ggplot(loadings) +
  aes(PC1, reorder(rownames(loadings),PC1), fill=PC1) +
  geom_col()
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

Let's make a nicer score plot with ggplot. Again, I need a data.frame with all the stuff I want (PC results and candy data) for my plot as input. 

```{r}
pc.results <- cbind(candy,pca$x)

ggplot(pc.results) +
  aes(PC1, PC2, label=rownames(pc.results)) +
  geom_point(col=my_cols) +
  geom_text_repel(col=my_cols) +
  labs(title="Candy Space via PCA")
```


> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Original variables that are picked up strongly by PC1 in the positive direction include fruity, hard, and pluribus. This is a bag/box of hard fruity candy that comes with multiple candies. Yes, this makes sense to me as shown clearly by the barplot. 